---
title: "A guide to mrf2d"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{A guide to mrf2d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mrf2d)
```

# Introduction

## A quick introduction to Markov Random Fields

 Markov Random Fields are probabilistic models used for modeling processes with local dependence. This dependence can be described by a graph $\mathcal{G} = (\mathcal{V}, \mathcal{N})$, where $\mathcal{V}$ is a set of vertices and $\mathcal{N}$ a neighborhood system.

$$ P_\theta(\mathbf{Z} = \mathbf{z}) = \frac{1}{\zeta_\theta} \exp \left( \sum_{q \in Q} g_q(\mathbf z_q, \theta) \right) $$

## A spatially stationary MRF model for 2-dimensional lattices

## The `mrf2d` package

# The `mrfi` class and parameter arrays

## Creating `mrfi` objects

## Parameter restriction families

### family `'onepar'`

### family `'oneeach'`

### family `'absdif'`

### family `'dif'`

### family `'free'`

# Parameter estimation on Markov Random Fields

## Pseudo-Likelihood

## Stochastic Approximation

# Hidden Markov Random Fields

## Corrupted discrete image model

## Gaussian Mixtures
